---
title: "cs-dmri vs MRtrix gold-standard fixel comparison"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
params:
  log_path:  "~/cs-bench-csdsi/focused/gold/run.log"
  gold_root: "~/cs-bench-csdsi/focused/gold"
  read_voxels: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5.2, dpi = 130)
suppressPackageStartupMessages({
  library(readr); library(dplyr); library(tidyr); library(stringr); library(purrr)
  library(forcats); library(ggplot2); library(scales); library(patchwork); library(knitr)
})

log_path  <- normalizePath(params$log_path,  mustWork = FALSE)
gold_root <- normalizePath(params$gold_root, mustWork = FALSE)

cfg_levels <- c("l2_baseline", "l1_ratio_1e-3",
                "l1_pathbic_n20", "l1_pathbic_ro4", "l1_pathbic_nn")
cfg_colors <- c("l2_baseline"    = "#888888",
                "l1_ratio_1e-3"  = "#d95f02",
                "l1_pathbic_n20" = "#1b9e77",
                "l1_pathbic_ro4" = "#7570b3",
                "l1_pathbic_nn"  = "#e7298a")
tag_colors <- c(direct = "#ca0020", roundtrip = "#0571b0")

theme_g <- theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right",
        panel.grid.minor = element_blank())
```

# What is being compared, and why

We are stress-testing **cs-dmri** — a compressed-sensing reconstruction that
fits a SHORE-style basis to a sparse HARDI / multi-shell DWI — against
the canonical **MRtrix3 MSMT-CSD** pipeline that qsirecon ships
(`mrtrix_multishell_msmt_noACT`) run on a high-quality reference DWI (ABCD).
The two tests answer two distinct questions:

1.  **`cs_direct_vs_gold` — does cs-odf's HASC55 fit recover the same fixels
    that MRtrix would have produced from the *real* ABCD DWI?**
    The HASC55 sequence is a sparsely sampled q-space scheme; cs's job is to
    fit a model that captures enough of the fiber geometry to be useful.
    "Useful" here is defined externally: peaks that line up, in 3-D, with what
    MRtrix MSMT-CSD finds in the dense reference acquisition. Differences
    here are *attribution-ambiguous* — they could be cs's signal model, cs's
    fixel extraction (peak finder + thresholding), or a real limitation of the
    HASC q-space coverage.

2.  **`cs_roundtrip_vs_gold` — does cs-predicted ABCD signal, fed through the
    *same* MRtrix pipeline, recover gold?**
    cs-fit on HASC → cs-synth onto ABCD's gradient table → MRtrix MSMT-CSD →
    fixels. Both sides of this comparison go through identical MRtrix code,
    so any "fixel extraction style" difference is cancelled out — what
    remains is purely *signal-prediction quality*. If `roundtrip ≈ direct`,
    cs's signal model is the bottleneck. If `roundtrip ≪ direct`, cs's
    in-house fixel extraction is contributing the bulk of the error.

A perfectly tuned cs-dmri would land *both* near zero discrepancy.
A reasonable one lands `roundtrip` clearly tighter than `direct`, and both
inside the deep-WM noise floor of MRtrix-vs-MRtrix.

## Per-metric reading guide

| Metric (per voxel)            | Lower is better? | What surprising values mean |
|-------------------------------|------------------|-----------------------------|
| `mean_match_angle_deg`        | yes              | >20° → fibers really are pointing differently; 30° is the cap |
| `top1_match`                  | no (1 = best)    | <0.5 → primary direction disagrees in the majority of voxels |
| `n_matched`                   | (paired)         | well below `min(n_a, n_b)` → each side is finding fixels the other isn't |
| `dice` / `jaccard` (presence) | no (1 = best)    | low + low angle → spurious extra peaks; low + high angle → real disagreement |
| `coherence_index`             | no (higher = aligned with neighbors) | drops < 0.10 → fixel field is noise |
| `dpf_amplitude_diff_mean`     | yes (signed)     | systematic bias = scaling difference in normalization |

# Data ingestion

We pull cell-level summary scalars straight from the gold-sweep stdout log,
and (optionally) read per-voxel NIfTI maps from each comparison cell.

```{r load_log}
parse_gold_log <- function(path) {
  lines <- readLines(path)
  pair_re <- "^===== pair: (\\S+) +<-> +(\\S+)"
  done_re <- "^\\[(cs_direct_vs_gold|cs_roundtrip_vs_gold)/([^/]+)/([^]]+)\\] done"

  cur <- list(); rows <- list(); pair_hasc <- NA_character_; pair_abcd <- NA_character_
  for (ln in lines) {
    m <- regmatches(ln, regexec(pair_re, ln))[[1]]
    if (length(m) == 3) { pair_hasc <- m[2]; pair_abcd <- m[3]; next }

    if (startsWith(ln, "primary_metric:"))     cur$primary_metric    <- trimws(sub("^primary_metric:", "", ln))
    if (startsWith(ln, "coherence_angle_deg:"))cur$coherence_angle   <- as.numeric(sub("^coherence_angle_deg:", "", ln))
    if (startsWith(ln, "match_angle_deg:"))    cur$match_angle       <- as.numeric(sub("^match_angle_deg:", "", ln))
    if (startsWith(ln, "voxels: a=")) {
      kv <- str_match(ln, "a=(\\d+), b=(\\d+), intersection=(\\d+)")[1, ]
      cur$voxels_a <- as.numeric(kv[2]); cur$voxels_b <- as.numeric(kv[3]); cur$voxels_isect <- as.numeric(kv[4])
    }
    if (startsWith(ln, "fixels: a=")) {
      kv <- str_match(ln, "a=(\\d+), b=(\\d+)")[1, ]
      cur$fixels_a <- as.numeric(kv[2]); cur$fixels_b <- as.numeric(kv[3])
    }
    if (startsWith(ln, "matched: mutual=")) {
      kv <- str_match(ln, "mutual=(\\d+), unmatched_a=(\\d+), unmatched_b=(\\d+)")[1, ]
      cur$mutual <- as.numeric(kv[2]); cur$unmatched_a <- as.numeric(kv[3]); cur$unmatched_b <- as.numeric(kv[4])
    }
    if (startsWith(ln, "mean_match_angle_deg:"))
      cur$mean_match_angle <- as.numeric(sub("^mean_match_angle_deg:", "", ln))
    if (startsWith(ln, "coherence_index: a=")) {
      kv <- str_match(ln, "a=([\\d\\.]+), b=([\\d\\.]+)")[1, ]
      cur$coh_a <- as.numeric(kv[2]); cur$coh_b <- as.numeric(kv[3])
    }
    if (startsWith(ln, "shared_dpf_keys:"))
      cur$shared_dpf_keys <- trimws(sub("^shared_dpf_keys:", "", ln))

    md <- regmatches(ln, regexec(done_re, ln))[[1]]
    if (length(md) == 4) {
      tag    <- sub("cs_", "", sub("_vs_gold", "", md[2]))   # "direct" | "roundtrip"
      pair   <- md[3]
      config <- md[4]
      rows[[length(rows) + 1]] <- modifyList(
        list(pair = pair, hasc = pair_hasc, abcd = pair_abcd, config = config, tag = tag),
        cur
      )
      cur <- list()
    }
  }
  bind_rows(rows)
}

scalars <- parse_gold_log(log_path) |>
  mutate(
    config       = factor(config, levels = cfg_levels),
    tag          = factor(tag,    levels = c("direct", "roundtrip")),
    subject      = str_extract(pair, "sub-[A-Za-z0-9]+"),
    matched_frac = mutual / pmin(fixels_a, fixels_b),
    a_extra      = unmatched_a / fixels_a,
    b_missed     = unmatched_b / fixels_b
  )
stopifnot(nrow(scalars) > 0)
```

```{r data_glance}
cat(sprintf("Cells parsed: %d  (%d pairs × %d configs × %d tags)\n",
            nrow(scalars),
            length(unique(scalars$pair)),
            length(unique(scalars$config)),
            length(unique(scalars$tag))))
```

# Per-cell summary

```{r per_cell_table}
scalars |>
  arrange(subject, config, tag) |>
  transmute(
    subject, config, tag,
    mean_angle_deg = round(mean_match_angle, 2),
    matched_frac   = round(matched_frac, 3),
    a_extra        = round(a_extra, 3),
    b_missed       = round(b_missed, 3),
    coh_a          = round(coh_a, 3),
    coh_b          = round(coh_b, 3),
    shared_dpf_keys
  )
```

# Per-config aggregates (median across subjects)

The headline numbers — what to look at first.

```{r per_config_table}
agg <- scalars |>
  group_by(config, tag) |>
  summarise(
    n              = n(),
    mean_angle_med = median(mean_match_angle),
    matched_frac   = median(matched_frac),
    a_extra        = median(a_extra),
    b_missed       = median(b_missed),
    coh_a          = median(coh_a),
    coh_b          = median(coh_b),
    .groups = "drop"
  ) |>
  arrange(config, tag)

agg |>
  mutate(across(where(is.numeric) & !n, ~ round(.x, 3))) |>
  kable(caption = "Medians across subjects, per (config, tag).")
```

The `direct` − `roundtrip` gap, at a glance:

```{r angle_gap_table}
scalars |>
  select(subject, config, tag, mean_match_angle) |>
  pivot_wider(names_from = tag, values_from = mean_match_angle) |>
  mutate(
    direct    = round(direct, 2),
    roundtrip = round(roundtrip, 2),
    gap       = round(direct - roundtrip, 2)
  ) |>
  arrange(config, subject) |>
  kable(caption = "Per-cell angle (deg). gap > 0 means roundtrip is tighter — cs's in-house fixel extraction is the larger source of error.")
```

# Plots

## 1. Mean match angle by config (direct vs roundtrip)

```{r p_angle_box}
p_angle <- ggplot(scalars, aes(fct_rev(config), mean_match_angle, fill = tag)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4, position = position_dodge(0.7), width = 0.6) +
  geom_point(aes(color = tag, group = tag), size = 2.6, alpha = 0.9,
             position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.7)) +
  scale_fill_manual(values  = tag_colors) +
  scale_color_manual(values = tag_colors) +
  coord_flip() +
  labs(title = "Mean match angle by config",
       subtitle = "Lower = closer to gold. Boxes pool 4 subjects per config × tag.",
       x = NULL, y = "mean match angle (deg)",
       fill = "test", color = "test") +
  theme_g
p_angle
```

## 2. Direct vs roundtrip, paired by subject

Each line connects the two tests for one subject × config. A line that
**slopes down** to roundtrip means cs's predicted signal is tighter than
its in-house fixels — the prediction is the better story, the local fixel
extractor is the weaker one. A flat line means the two stories agree:
whatever discrepancy exists is in the signal model, not the fixel routine.

```{r p_paired}
paired <- scalars |>
  select(subject, config, tag, mean_match_angle) |>
  pivot_wider(names_from = tag, values_from = mean_match_angle)

paired_long <- scalars |>
  select(subject, config, tag, mean_match_angle)

p_pair <- ggplot(paired_long, aes(tag, mean_match_angle)) +
  geom_line(aes(group = interaction(subject, config), color = subject),
            alpha = 0.85, linewidth = 0.7) +
  geom_point(aes(color = subject, shape = subject), size = 2.6) +
  facet_wrap(~ config, nrow = 1) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Per-subject paired direct vs roundtrip",
       subtitle = "Down-sloping line: cs prediction better than cs's own fixels. Flat: signal-model bound.",
       x = NULL, y = "mean match angle (deg)") +
  theme_g +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
p_pair
```

## 3. Detection accounting — matched vs spurious vs missed

`a_extra` is the fraction of side-A fixels with no acceptable match on B
(false positives, from gold's perspective); `b_missed` is the fraction of
gold-side fixels cs failed to find (false negatives).

```{r p_counts}
counts_long <- scalars |>
  select(subject, config, tag, matched_frac, a_extra, b_missed) |>
  pivot_longer(c(matched_frac, a_extra, b_missed),
               names_to = "kind", values_to = "frac") |>
  mutate(kind = factor(kind,
                       levels = c("matched_frac", "a_extra", "b_missed"),
                       labels = c("matched (mutual / min(a,b))",
                                  "a extra (cs has, gold doesn't)",
                                  "b missed (gold has, cs doesn't)")))

p_counts <- ggplot(counts_long, aes(config, frac, fill = tag)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4, position = position_dodge(0.7), width = 0.6) +
  geom_point(aes(color = tag, group = tag), size = 2.2, alpha = 0.9,
             position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.7)) +
  facet_wrap(~ kind, scales = "free_y") +
  scale_fill_manual(values = tag_colors) +
  scale_color_manual(values = tag_colors) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Fixel detection accounting",
       x = NULL, y = "fraction", fill = "test", color = "test") +
  theme_g +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
p_counts
```

## 4. Fixel-field coherence — cs vs gold

`coherence_index` measures how aligned a voxel's fixels are with its
neighbors. The gold side is identical across configs (same MRtrix run);
cs's coherence varies by regularization. A cs run with coherence
*higher* than gold can mean *over-smoothed*; lower can mean *noisy*.

```{r p_coherence}
coh <- scalars |>
  filter(tag == "direct") |>     # gold value is the same in both rows; pick one
  select(subject, config, coh_a, coh_b)

p_coh <- ggplot(coh, aes(config, coh_a, color = subject, shape = subject)) +
  geom_hline(aes(yintercept = coh_b), data = coh,
             linetype = "dashed", color = "grey50", alpha = 0.5) +
  geom_point(size = 3.2, alpha = 0.9) +
  geom_text(aes(y = coh_b, label = sprintf("gold = %.2f", coh_b)),
            color = "grey40", size = 2.8, hjust = 0, nudge_x = 0.15,
            data = coh |> distinct(subject, coh_b) |> mutate(config = factor(cfg_levels[1], levels = cfg_levels))) +
  scale_color_brewer(palette = "Dark2") +
  facet_wrap(~ subject, nrow = 1) +
  labs(title = "cs-side coherence vs MRtrix gold",
       subtitle = "Dashed = gold (subject-specific). cs above gold ≈ over-smoothed; below ≈ noisier.",
       x = NULL, y = "coherence_index") +
  theme_g +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
p_coh
```

# Per-voxel distributions

```{r read_voxels, eval=params$read_voxels}
have_rnifti <- requireNamespace("RNifti", quietly = TRUE)
if (!have_rnifti) {
  message("RNifti not installed; skipping per-voxel section. Install with: install.packages('RNifti')")
}
```

```{r voxel_load, eval=params$read_voxels && exists("have_rnifti") && have_rnifti}
read_dpv <- function(pair, config, tag_full, name) {
  p <- file.path(gold_root, pair, config, tag_full, paste0(name, ".nii.gz"))
  if (!file.exists(p)) return(NULL)
  arr <- as.numeric(RNifti::readNifti(p))
  arr[is.finite(arr) & arr > 0]
}

cells <- scalars |> distinct(pair, subject, config, tag) |>
  mutate(tag_full = paste0("cs_", tag, "_vs_gold"))

vox_angle <- cells |>
  mutate(values = pmap(list(pair, config, tag_full),
                       \(p, c, t) read_dpv(p, as.character(c), t, "mean_match_angle_deg"))) |>
  filter(map_lgl(values, ~ !is.null(.x) && length(.x) > 0)) |>
  unnest(values) |>
  rename(angle = values)

vox_top1 <- cells |>
  mutate(values = pmap(list(pair, config, tag_full),
                       \(p, c, t) {
                         v <- read_dpv(p, as.character(c), t, "top1_match")
                         if (is.null(v)) NULL else v
                       })) |>
  filter(map_lgl(values, ~ !is.null(.x) && length(.x) > 0)) |>
  unnest(values) |>
  rename(top1 = values)

vox_amp <- cells |>
  mutate(values = pmap(list(pair, config, tag_full),
                       \(p, c, t) {
                         f <- file.path(gold_root, p, as.character(c), t, "dpf_amplitude_diff_mean.nii.gz")
                         if (!file.exists(f)) return(NULL)
                         arr <- as.numeric(RNifti::readNifti(f))
                         arr[is.finite(arr)]
                       })) |>
  filter(map_lgl(values, ~ !is.null(.x) && length(.x) > 0)) |>
  unnest(values) |>
  rename(amp_diff = values)
```

## Voxel-level mean match angle

This is the most direct view of *where the test fails*. The boxes are over
voxels (not subjects). Tails reaching the 30° cap = voxels that picked up
a match only because the threshold allowed it; mass at low angles = real
agreement.

```{r p_vox_angle, eval=params$read_voxels && exists("vox_angle") && nrow(vox_angle) > 0}
p_vox_angle <- ggplot(vox_angle, aes(fct_rev(config), angle, fill = tag)) +
  geom_violin(alpha = 0.45, position = position_dodge(0.7), width = 0.9, scale = "width", trim = TRUE) +
  geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.7,
               position = position_dodge(0.7)) +
  scale_fill_manual(values = tag_colors) +
  coord_flip() +
  labs(title = "Per-voxel mean match angle distribution",
       subtitle = "Pooled across subjects. Mass at low values = real recovery; mass at the cap = forced matches.",
       x = NULL, y = "mean match angle (deg)", fill = "test") +
  theme_g
p_vox_angle
```

## Voxel-level top-1 match (primary direction agreement)

`top1_match = 1` when the largest fixel on side A is matched to the largest
on side B and falls within the match-angle threshold. This is the
"single-fiber WM voxel" sanity check — in deep WM these should be near 1
across most voxels.

```{r p_vox_top1, eval=params$read_voxels && exists("vox_top1") && nrow(vox_top1) > 0}
p_vox_top1 <- ggplot(vox_top1, aes(top1, fill = tag, color = tag)) +
  stat_ecdf(geom = "step", linewidth = 0.8, alpha = 0.85) +
  facet_wrap(~ config, nrow = 1) +
  scale_color_manual(values = tag_colors) +
  scale_fill_manual(values  = tag_colors) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(title = "ECDF of top-1 match (per voxel, pooled across subjects)",
       subtitle = "Curve hugging the right = primary direction usually agrees.",
       x = "top1_match", y = "F(x)", color = "test", fill = "test") +
  theme_g
p_vox_top1
```

## Amplitude bias (signed)

`dpf_amplitude_diff_mean` is signed: positive = cs side has larger matched
fixel amplitude than gold. A clean run sits on zero with symmetric tails.
A consistently shifted distribution is a *normalization* problem (e.g.
`mtnormalise` not applied, basis scaling drift), not a directional one.

```{r p_amp, eval=params$read_voxels && exists("vox_amp") && nrow(vox_amp) > 0}
amp_clip <- vox_amp |>
  group_by(config, tag) |>
  mutate(amp_diff = pmin(pmax(amp_diff, quantile(amp_diff, .005)),
                                          quantile(amp_diff, .995))) |>
  ungroup()

p_amp <- ggplot(amp_clip, aes(amp_diff, fill = tag, color = tag)) +
  geom_density(alpha = 0.35) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~ config, scales = "free_y") +
  scale_color_manual(values = tag_colors) +
  scale_fill_manual(values  = tag_colors) +
  labs(title = "Per-voxel mean matched-fixel amplitude difference (cs − gold)",
       subtitle = "Density, 0.5–99.5% clipped per panel. Centered = no scale bias; shifted = normalization issue.",
       x = "amplitude diff (cs − gold)", y = "density", color = "test", fill = "test") +
  theme_g
p_amp
```

# Practical takeaways

```{r takeaways}
best_direct    <- agg |> filter(tag == "direct")    |> slice_min(mean_angle_med, n = 1)
best_roundtrip <- agg |> filter(tag == "roundtrip") |> slice_min(mean_angle_med, n = 1)
worst_direct   <- agg |> filter(tag == "direct")    |> slice_max(mean_angle_med, n = 1)

cat(sprintf("Best on direct:    %-15s  median angle %5.2f°\n",
            best_direct$config, best_direct$mean_angle_med))
cat(sprintf("Best on roundtrip: %-15s  median angle %5.2f°\n",
            best_roundtrip$config, best_roundtrip$mean_angle_med))
cat(sprintf("Worst on direct:   %-15s  median angle %5.2f°\n",
            worst_direct$config, worst_direct$mean_angle_med))
```

**How to read the headline numbers.** A median match angle of \~12–15° is
consistent with what MRtrix-vs-MRtrix scatter looks like in clean WM at
this match threshold (30°); the floor is set by gold's own fixel-detection
noise, not by cs. Configs that drag above \~16° on `direct` *and* fail to
recover on `roundtrip` are flagging a real signal-model problem (e.g. the
non-negative path-bic config hitting a constrained optimum that distorts
predicted signal). Configs that improve on `roundtrip` but not `direct`
are flagging a fixel-extraction style issue — cs's own peak finder /
threshold differs from MRtrix's.

**What the gap means in practice.**

-   `gap = direct − roundtrip > 0` and large (≳ 1°): swap cs's in-house
    fixel extractor for fod2fixel-on-cs-coeffs in downstream analyses.
    The signal is fine; the fixel routine is the cost.
-   `gap ≈ 0`: cs's fixel routine is already fine; further angle reduction
    requires a better signal model (denser q-space, different basis, less
    aggressive regularization).
-   `gap < 0` (rare): the predicted signal *adds* error — the regularizer
    is too aggressive and is biasing the predicted DWI. Inspect that
    config's `cross_r2_p50` from the bench summary.

# Render

```bash
Rscript -e "rmarkdown::render('cs-dmri/scripts/gold_compare.Rmd', \\
  params = list(log_path = '<path/to/gold sweep stdout>', \\
                gold_root = '~/cs-bench-csdsi/focused/gold', \\
                read_voxels = TRUE), \\
  output_dir = '~/cs-bench-csdsi/focused/gold')"
```
